---
title: "Separating Gaussians"
subtitle: "An explainer"
author: "Dan Killian"
date: today
toc: true
toc-depth: 3
number-sections: false
format:
  html:
    code-fold: true
    page-layout: full
execute:
  keep-md: true
editor: visual
reference-location: margin
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../prep (May 2025).r")
```

## The Problem

We observe a sample of heights and know that it was drawn from a population composed of two groups — female and male — each with its own distribution of heights. The question is: given only a height measurement, what is the probability that the observation came from each group?

We model each group's heights as Gaussian, so the full population is a **Gaussian mixture**:

$$f(x) = \pi_f \cdot \phi(x; \mu_f, \sigma_f) + \pi_m \cdot \phi(x; \mu_m, \sigma_m)$$

where $\phi(x; \mu, \sigma)$ is the Gaussian density and $\pi_k$ is the **mixture weight** for group $k$ — the prior probability that a randomly drawn person belongs to that group.

### The data

```{r}
library(NHANES)

heights <- NHANES |>
    filter(Age >= 20, !is.na(Height), !is.na(Gender)) |>
    select(sex=Gender, age=Age, cm=Height) %>%
    mutate(inches=round(cm/2.54,0),
           ft=floor(inches/12),
           rem=round(inches%% 12,1),
           feet=round(inches/12,1),
           feet_lab = paste(ft, "'", rem, "\"", sep=""))
```


```{r}
library(flextable)

parms <- heights |>
  group_by(sex) |>
  summarise(
    n       = n(),
    mean_cm = mean(cm, na.rm = TRUE),
    sd_cm   = sd(cm,   na.rm = TRUE)
  ) 

prior_f <- parms$n[1] / sum(parms$n)
prior_m <- parms$n[2] / sum(parms$n)

parms_flx <- parms |>
    mutate(weight = n / sum(n)) |>
    select(sex, n, weight, mean_cm, sd_cm) |>
    flextable() |>
    colformat_double(j = "weight", digits = 3) |>
    colformat_double(j = c("mean_cm", "sd_cm"), digits = 1)

parms_flx
```

```{r}
ggplot(heights, aes(x=cm, color=sex, fill=sex)) +
    geom_density(alpha=.2) +
    scale_fill_viridis_d() +
    scale_color_viridis_d() +
    scale_y_continuous(breaks=NULL) +
    labs(y="",
         title="Distribution of Heights")
```


### Mixture weights

The mixture weight $\pi_k$ is the share of the total sample belonging to group $k$:

$$\pi_k = \frac{n_k}{N}, \qquad N = \sum_k n_k$$

For this dataset, $\pi_f =$ `r round(parms$weight[1], 3)` and $\pi_m =$ `r round(parms$weight[2], 3)`. These reflect the sample composition and serve as **prior probabilities** before observing any height.

Without mixture weights, we would implicitly treat each group as equally probable — equivalent to setting $\pi_f = \pi_m = 0.5$. When the groups differ in size, this introduces a systematic bias in classification.

```{r}
heights <- heights |>
    mutate(lik_female = dnorm(cm,
                            mean=parms$mean_cm[1],
                            sd=parms$sd_cm[1]),
           lik_male = dnorm(cm,
                            mean = parms$mean_cm[2],
                            sd = parms$sd_cm[2]),
           p_female   = (lik_female * prior_f) / 
               (lik_female * prior_f + lik_male * prior_m),
           p_male = ( lik_male * prior_m ) / 
               ( (lik_male * prior_m) + (lik_female * prior_f) )
    )

heights <- heights %>%
    group_by(sex) %>%
    mutate(percentile=cume_dist(cm))
           
```


```{r}
library(ggridges)

ggplot(heights, aes(x = cm, y = sex)) +
  geom_density_ridges()
```

```{r}
ggplot(heights, aes(x = cm, fill = sex)) +
  geom_dotplot(binwidth = 2, dotsize=.4) +
    scale_x_continuous(breaks=seq(100, 220, 20)) +
  facet_wrap(~ sex, ncol = 1) + 
    faceted
```

```{r}
ggplot(heights, aes(cm)) +
    geom_point(aes(y=p_male),
               size=.5,
               color="dodgerblue2") +
    geom_point(aes(y=p_female),
               size=.5,
               color="indianred") +
    scale_x_continuous(limits=c(135,205),
                       breaks=seq(140,200,10))
```








## The Geometry

## The Algorithm

## Simulation
