---
title: "The Law of Iterated Expectations"
subtitle: "Also known as the tower property or Adam's law"
author: "Dan Killian"
date: today
toc: true
toc-depth: 3
number-sections: false
format:
  html:
    code-fold: true
    page-layout: full
execute:
  keep-md: true
editor: visual
reference-location: margin
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../prep (May 2025).R")
```

## The law

The **law of iterated expectations** (also called the tower property or Adam's law) states:

$$E[X] = E\big[E[X \mid Y]\big]$$

In words: the overall average of X equals the weighted average of the group-specific averages of X, where the weights are the group sizes.

This sounds obvious once you see it — but it's worth being precise about what "weighted average" means here, because a common mistake is to average group means without accounting for group size.

---

## The setup

We have a sample of 1,000 people. For each person we observe:

- **Y** — their highest level of education (High School, College, or Graduate)  
- **X** — their annual income (in USD)

Income tends to be higher for more educated workers, and the three groups are different sizes.

```{r}
set.seed(5842)

n <- 1000

education <- sample(
  c("High School", "College", "Graduate"),
  size    = n,
  prob    = c(0.40, 0.45, 0.15),
  replace = TRUE
)

income <- case_when(
  education == "High School" ~ rnorm(n, mean = 40000, sd = 8000),
  education == "College"     ~ rnorm(n, mean = 65000, sd = 12000),
  education == "Graduate"    ~ rnorm(n, mean = 95000, sd = 20000)
)

df <- data.frame(education, income)
head(df)
```

---

## Step 1: The overall mean E[X]

The overall mean income is straightforward:

```{r}
EX <- mean(df$income)
cat("E[X] = overall mean income: $", round(EX, 0))
```

---

## Step 2: The conditional means E[X | Y]

Now compute the mean income *within each education group*. This gives us E[X | Y] — a function of Y that assigns each person their group's average income.

```{r}
library(tidyverse)

group_stats <- df |>
  summarise(
    n       = n(),
    mean    = mean(income),
    .by     = education
  ) |>
  mutate(
    prop    = n / sum(n),
    education = factor(education, levels = c("High School", "College", "Graduate"))
  ) |>
  arrange(education)

group_stats
```

```{r}
#| echo: false
ggplot(group_stats, aes(x = education, y = mean)) +
  geom_col(fill = medium_blue, width = 0.5) +
  geom_hline(yintercept = EX, color = usaid_red, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 0.6, y = EX + 1200,
           label = paste0("E[X] = $", round(EX, 0)),
           color = usaid_red, hjust = 0, size = 4) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Mean income by education group",
    subtitle = "Dashed line = overall mean",
    x = NULL, y = "Mean income"
  )
```

---

## Step 3: E[E[X | Y]] — iterating the expectation

To get E[E[X | Y]], we take the expectation of the conditional mean over the distribution of Y. Since Y is discrete, this is a weighted average of group means, weighted by group proportions:

$$E\big[E[X \mid Y]\big] = \sum_y E[X \mid Y = y] \cdot P(Y = y)$$

```{r}
E_EXY <- sum(group_stats$mean * group_stats$prop)
cat("E[E[X|Y]] = $", round(E_EXY, 0))
```

---

## Step 4: Verify the law

```{r}
cat(
  "E[X]        = $", round(EX,    0), "\n",
  "E[E[X|Y]]  = $", round(E_EXY, 0), "\n",
  "Difference  =  ", round(EX - E_EXY, 6)
)
```

They're equal (up to simulation noise). The law holds.

---

## The key pitfall: unweighted averages

A common mistake is to average group means *without* weighting by group size:

```{r}
unweighted_mean <- mean(group_stats$mean)

cat(
  "Weighted avg of group means   (correct): $", round(E_EXY,           0), "\n",
  "Unweighted avg of group means (wrong):   $", round(unweighted_mean, 0)
)
```

These differ because the three groups are not the same size. High School (40%) and College (45%) workers outnumber Graduate (15%) workers. The unweighted average gives Graduate workers too much influence, pulling the result upward.

```{r}
#| echo: false
comparison <- data.frame(
  method = c("E[X]\n(true mean)", "E[E[X|Y]]\n(weighted)", "Simple avg\nof group means"),
  value  = c(EX, E_EXY, unweighted_mean),
  correct = c(TRUE, TRUE, FALSE)
)

ggplot(comparison, aes(x = method, y = value, fill = correct)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0("$", round(value, 0))),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("TRUE" = medium_blue, "FALSE" = usaid_red),
                    labels = c("TRUE" = "Correct", "FALSE" = "Incorrect"),
                    name = NULL) +
  scale_y_continuous(labels = scales::dollar, limits = c(0, 75000)) +
  labs(
    title = "Weighted vs. unweighted average of group means",
    subtitle = "Only the weighted average equals E[X]",
    x = NULL, y = "Estimated mean income"
  )
```

The law of iterated expectations is precise: the weights must come from the *marginal distribution of Y* — i.e., the actual proportion of people in each group.

---

## Why it works: decomposing the sum

We can see the arithmetic directly. The overall mean is just the total income divided by n:

$$E[X] = \frac{\sum_i x_i}{n} = \frac{\sum_y \left( \sum_{i: Y_i = y} x_i \right)}{n}$$

Within each group, the sum of incomes equals the group mean times the group size:

$$= \frac{\sum_y \bar{x}_y \cdot n_y}{n} = \sum_y \bar{x}_y \cdot \frac{n_y}{n} = \sum_y E[X \mid Y = y] \cdot P(Y = y)$$

That last expression is exactly $$E[E[X \mid Y]]$$.

```{r}
# Verify the arithmetic directly
group_stats |>
  mutate(contribution = mean * prop) |>
  summarise(
    `E[E[X|Y]]` = sum(contribution)
  )
```

---

## Summary

The law of iterated expectations says the overall mean is the probability-weighted average of group means:

$$E[X] = E\big[E[X \mid Y]\big] = \sum_y E[X \mid Y = y] \cdot P(Y = y)$$

Three names, one idea:

| Name | Context |
|------|---------|
| Law of iterated expectations | Probability / statistics |
| Tower property | Measure-theoretic probability |
| Adam's law | Econometrics / applied statistics |

The key practical lesson: when averaging subgroup means to recover an overall mean, **weight by group size**. An unweighted average of group means is not the same as the population mean unless all groups are the same size.
